pirouette: Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-03-15 article.tex
Process feedback @Giappo, thanks @Giappo

\giovanni{
  This is a slightly more formal version. 
  I personally find it clearer. 
  If you don't like it we can use version 2 (see above in the source). 
  In case you really like it, instead, I can try to make the rest a bit more similar to this.
}

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the given phylogeny. 
To do so, we must specify an alignment model $\mathit{A}$ which consists of: a DNA root sequence $\mathit{r_{s}}$, the DNA mutation rate $\mathit{m_{r}}$, a
site model $\mathit{s_{m}}$ (in this context: a nucleotide substitution model) and a clock model $\mathit{c_{m}}$. 
In this example, the DNA root sequence consists in four blocks of 250 nucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.
\iffalse
\giovanni{this looks unnecessary to me}
Because the crown age of the phylogeny is 10 time units, each nucleotide will
have one (possibly-silent) mutation on average.
\fi
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A Jukes-Cantor site model,
assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all species are equal and constant.

@@

\iffalse
\giovanni{I think we are providing too many information. Let's try to keep the main text as dry as possible.}
One the hardware we used, it will take approximately 10 minutes
to obtain the error distribution returned by \verb;pir_run;.
This error distribution is a data frame in the non-tidy short form.
Instead of converting it to tidy data and plotting it,
\verb;pirouette; has a plotting function that does exactly that:
\fi

@@

\iffalse
This error distribution has little relevance on its own.
\giovanni{maybe it's better to explain later the multiple tree argument}
It only gives a baseline error made by
the Bayesian inference. We use 'a', because we have used only one Yule
tree. To obtain the baseline error, one should simulate multiple Yule
trees and merge their error distributions.
\fi

\iffalse
For this error distribution to have more relevance it needs to be compared
to an error distribution derived from a phylogeny 
derived from a non-standard speciation process.
\fi

@@

\iffalse
By doing so, this error distribution can serve as a control,
as it can be seen as the baseline error caused by stochasticity.
In this example, that non-standard speciation process must 
be closest related to a Yule process for the comparison to make sense \giovanni{why?}.
Would the two error distributions be similar, the non-standard speciation 
process is of little influence.
Because measuring the baseline error is so important,
we made it an inherent part of \verb;pirouette;. 
We call it twinning and we use it in example research question 3.
\fi

@@

\iffalse
There are more ways to pick a standard inference model. In the
previous example, we picked the known generative inference model.
\fi
In the previous example, we used a known generative tree prior to match with the inference model.
However, a novel tree prior, by definition, is non-standard, thus cannot be part of a standard inference model (i.e. an inference model using a standard tree prior). 
One option is to choose to hand-pick a related 
standard tree prior, as done in ERCs 3 and 5, and combine it with 
a known generative alignment model.
An alternative is to let multiple candidate inference models compete 
and pick the best one, based on the highest evidence (aka marginal likelihood):
when an alignment is simulated from the given tree, one can estimate the
evidence for each inference model given that 
alignment \richel{or the other way around?}.
The benefit of this approach is that this is a commonly used procedure,
which picks an inference model that is
best at explaining the alignment, where more complex inference models are
penalized for having more parameters.

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_3_errors.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, picking the best inference model versus the control.
    Here, the twin column shows the error BEAST2 makes starting from the original tree versus the twin. 
  }
  \label{fig:example_3}
  \giovanni{remove boxplots}
\end{figure}

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_4_errors.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative model is known, compared to candidates with similar inference models, in relationship to the background error.
  }
  \label{fig:example_4}
  \giovanni{Remove boxplots.}
\end{figure}

>>>>

\giovanni{This is a slightly more formal version}
\richel{As agreed, go ahead and expend this formalization}

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the given phylogeny. 
To do so, we must specify an alignment model $\mathit{A}$ which consists of: a DNA root sequence $\mathit{r_{s}}$, the DNA mutation rate $\mathit{m_{r}}$, a
site model $\mathit{s_{m}}$ (in this context: a nucleotide substitution model) and a clock model $\mathit{c_{m}}$. 
In this example, the DNA root sequence consists in four blocks of 250 nucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A Jukes-Cantor site model,
assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all species are equal and constant.

@@

\giovanni{I think we are providing too many information}
\richel{agreed, thanks for pushing back}

@@

@@

@@

In the previous example, we used a known generative tree prior to match with the inference model.
However, a novel tree prior, by definition, is non-standard, thus cannot be part of a standard inference model (i.e. an inference model using a standard tree prior). 
One option is to choose to hand-pick a related 
standard tree prior, as done in ERCs 3 and 5, and combine it with 
a known generative alignment model.
An alternative is to let multiple candidate inference models compete 
and pick the best one, based on the highest evidence (aka marginal likelihood):
when an alignment is simulated from the given tree, one can estimate the
evidence for each inference model given that 
alignment \richel{or the other way around?}.
The benefit of this approach is that this is a commonly used procedure,
which picks an inference model that is
best at explaining the alignment, where more complex inference models are
penalized for having more parameters.

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_3_errors.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, picking the best inference model versus the control.
    Here, the twin column shows the error BEAST2 makes starting from the original tree versus the twin. 
  }
  \label{fig:example_3}
\end{figure}

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_4_errors.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative model is known, compared to candidates with similar inference models, in relationship to the background error.
  }
  \label{fig:example_4}
\end{figure}
